\documentclass[11pt,a4paper]{article}

\input{preamble.tex}

\title{The \texttt{smc2fc} Framework \\
\large A Practical Reference for Sequential Monte Carlo$^2$ \\
Filtering and Stochastic Optimal Control}
\author{smc2fc development team}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent
This document is a practical reference for the \texttt{smc2fc} package
--- a JAX implementation of Sequential Monte Carlo$^2$ for Bayesian
filtering of state-space models with intractable likelihoods, paired
with a control-as-inference receding-horizon controller. It is
deliberately model-independent: the underlying SDE model and observation
likelihood are abstractions supplied by the user, and nothing in this
document depends on the specifics of any particular application.

The intended reader is a research engineer who needs to use, extend, or
tune the framework. The treatment is technical where it must be (the
Bayesian filtering recursion, the unbiased likelihood property, the
adaptive tempering schedule) and pedagogic where it helps (the
control-as-inference framing, the receding-horizon principle, the
soft-barrier feasibility argument). Two kinds of empirical findings are
documented: the JAX-native compile-once architecture (which speeds up
inference by an order of magnitude) and the result that NUTS, despite
being the textbook upgrade from HMC, is roughly 100$\times$ slower than
HMC inside this particular nested control flow. Both are written so
future agents do not repeat the dead-end.

The document deliberately omits speculation about algorithmic
alternatives the framework does not currently implement. The one open
direction it does cover --- improving the soft-HMC controller's wall-clock
without sacrificing control quality --- is the only bottleneck the
authors have hit empirically.
\end{abstract}

\tableofcontents
\newpage

\input{sections/01_overview.tex}
\input{sections/02_setting.tex}
\input{sections/03_particle_filter.tex}
\input{sections/04_ot_rescue.tex}
\input{sections/05_smc2.tex}
\input{sections/06_jax_native.tex}
\input{sections/07_gaussian_bridge.tex}
\input{sections/08_control_inference.tex}
\input{sections/09_mpc.tex}
\input{sections/10_config.tex}
\input{sections/11_future_work.tex}

\appendix
\input{sections/appendix_D_code_api.tex}

\end{document}
