\section{The FSA-v2 SDE model}\label{sec:fsa}

\subsection{State, drift, and diffusion}

Let $(B(t),\,F(t),\,A(t))$ denote the latent state of a single individual
at time $t \ge 0$, measured in days. The three components are interpreted as
\begin{itemize}
  \item $B(t) \in [0,1]$ -- a normalised \emph{fitness} variable; the
        Banister chronic compartment;
  \item $F(t) \in [0,\infty)$ -- an unnormalised \emph{fatigue} variable;
        the Banister acute compartment;
  \item $A(t) \in [0,\infty)$ -- the \emph{amplitude} of an autonomic /
        circadian Stuart--Landau oscillator.
\end{itemize}
The single exogenous control input is the (non-negative) training-stimulus
rate $\Phi(t) \in [0,\Phimax]$, expressed in arbitrary daily TRIMP-like
units. Throughout we work in It\^o form. The drift coefficients are
\begin{align}
  \mu(B,F) &= \mu_0 + \mu_B B - \mu_F F - \mu_{FF} F^2, \label{eq:fsa-mu} \\
  \dd B &= \bigl[\,\kappa_B(1+\varepsilon_A A)\,\Phi(t) - B/\tau_B\,\bigr]\,\dd t
            + \sigma_B \sqrt{B(1-B)}\,\dd W_B, \label{eq:fsa-B} \\
  \dd F &= \bigl[\,\kappa_F\,\Phi(t) - (1+\lambda_A A)/\tau_F \cdot F\,\bigr]\,\dd t
            + \sigma_F \sqrt{F}\,\dd W_F, \label{eq:fsa-F} \\
  \dd A &= \bigl[\,\mu(B,F)\,A - \eta\,A^3\,\bigr]\,\dd t
            + \sigma_A \sqrt{A}\,\dd W_A, \label{eq:fsa-A}
\end{align}
where $W_B,W_F,W_A$ are independent scalar Brownian motions. The diffusion
in \eqref{eq:fsa-B} is the Jacobi diffusion on $[0,1]$ (Forman and S\o rensen,
2008): the volatility vanishes at both boundaries $B=0$ and $B=1$, so the
state never leaves the unit interval. The diffusions in \eqref{eq:fsa-F}
and \eqref{eq:fsa-A} are square-root (CIR-type, after Cox, Ingersoll and
Ross, 1985): the volatility vanishes at the lower boundary, so $F$ and $A$
remain non-negative.

The drift terms have a natural physiological reading. Equation
\eqref{eq:fsa-B} is the standard Banister chronic equation
$\dot B = \kappa_B \Phi - B/\tau_B$, augmented by an autonomic-coupling
factor $(1+\varepsilon_A A)$: when the autonomic amplitude is high the
body adapts more efficiently from the same training stimulus. Equation
\eqref{eq:fsa-F} is the Banister acute equation, with the clearance rate
multiplied by $(1+\lambda_A A)$: a high autonomic amplitude clears fatigue
faster. Equation \eqref{eq:fsa-A} is a Stuart--Landau cubic, in which the
linear growth rate $\mu(B,F)$ depends on the slow Banister state. As
$\mu$ crosses zero the deterministic skeleton undergoes a Hopf-type
bifurcation: for $\mu<0$ the origin is stable, and for $\mu>0$ a stable
limit cycle of radius $\sqrt{\mu/\eta}$ appears. The quadratic
$-\mu_{FF} F^2$ term in \eqref{eq:fsa-mu} encodes \emph{overtraining
collapse}: when fatigue is excessive, the autonomic system loses its
ability to oscillate.

\subsection{Truth parameters}

Table~\ref{tab:fsa-truth} lists the truth parameters used throughout
these notes. They correspond to the canonical Banister time constants
$\tau_B = 42\,\mathrm{d}$, $\tau_F = 7\,\mathrm{d}$ and to a Stuart--Landau
amplitude with deterministic equilibrium $A^* \approx 1$ when
$\mu \approx 0.2$.

\begin{table}[h]
\centering
\small
\begin{tabular}{@{}llrl@{}}
\toprule
Group & Symbol & Value & Meaning \\
\midrule
Banister timescales       & $\tau_B$              & $42$       & chronic time constant (d) \\
                          & $\tau_F$              & $7$        & acute time constant (d) \\
Banister gains            & $\kappa_B$            & $0.012$    & $B$-gain per unit $\Phi$ \\
                          & $\kappa_F$            & $0.030$    & $F$-gain per unit $\Phi$ \\
$A$-coupling              & $\varepsilon_A$       & $0.40$     & boost to $B$-gain at $A=1$ \\
                          & $\lambda_A$           & $1.00$     & boost to $F$-clearance at $A=1$ \\
Stuart--Landau            & $\mu_0$               & $0.02$     & baseline growth rate \\
                          & $\mu_B$               & $0.30$     & fitness raises $\mu$ \\
                          & $\mu_F$               & $0.10$     & linear fatigue suppression \\
                          & $\mu_{FF}$            & $0.40$     & quadratic fatigue collapse \\
                          & $\eta$                & $0.20$     & cubic damping \\
Diffusion                 & $\sigma_B$            & $0.010$    & Jacobi noise scale \\
                          & $\sigma_F$            & $0.012$    & CIR noise scale on $F$ \\
                          & $\sigma_A$            & $0.020$    & CIR noise scale on $A$ \\
\bottomrule
\end{tabular}
\caption{Truth parameters of the FSA-v2 dynamics. Source:
\texttt{version\_2/models/fsa\_high\_res/\_dynamics.py}.}
\label{tab:fsa-truth}
\end{table}

\subsection{The training-stimulus envelope}

The control variable in the SDE is the instantaneous training-stimulus
rate $\Phi(t)$. In practice training is delivered in discrete daily
sessions concentrated in the morning, and the simulator expands a daily
schedule $(\Phi_d)_{d=1,\dots,N}$ into a high-resolution envelope at
fifteen-minute resolution as follows. Let $\tau_{\mathrm{burst}}=3\,\mathrm{h}$ be a
shape parameter, and let $[t_{\mathrm{wake}},t_{\mathrm{sleep}}] = [7,23]$ be the
daily wake window (in hours). The within-day envelope is
\begin{equation}
  e(t) \;\propto\;
    \begin{cases}
      (t-t_{\mathrm{wake}})\,\exp\!\bigl(-(t-t_{\mathrm{wake}})/\tau_{\mathrm{burst}}\bigr)
        & t_{\mathrm{wake}} \le t < t_{\mathrm{sleep}}, \\
      0  & \text{otherwise},
    \end{cases}
  \label{eq:phi-envelope}
\end{equation}
which is the standard Gamma$(k=2)$ shape, peaking at
$t-t_{\mathrm{wake}} = \tau_{\mathrm{burst}}$. The constant of proportionality
is fixed so that
\begin{equation}
  \int_{\text{day}} \Phi(t)\,\dd t \;=\; 24\,\Phi_d,
\end{equation}
i.e.~the daily integral of the within-day envelope is preserved. The slow
Banister equations \eqref{eq:fsa-B}--\eqref{eq:fsa-F} therefore see the
same daily load whether $\Phi$ is delivered as a constant or as a
morning-loaded burst, and the fast Stuart--Landau dynamics in $A$ pick up
the within-day shape.

\subsection{Four-channel observation model}

The state $(B,F,A)$ is not directly observed. Instead we receive four
discrete-time channels at the fifteen-minute grid:
\begin{align}
  \mathrm{HR}(t) &\sim
    \Ncal\!\bigl(\,\mathrm{HR}_{\mathrm{base}}
                  - \kappa_B^{\mathrm{HR}}\,B
                  + \alpha_A^{\mathrm{HR}}\,A
                  + \beta_C^{\mathrm{HR}}\,C(t),\;
                  \sigma_{\mathrm{HR}}^2\bigr),
    && \text{(sleep-gated)} \label{eq:obs-hr}\\
  P(\mathrm{sleep}(t)=1\mid A,C) &=
    \sigmoid\!\bigl(\,k_C\,C(t) + k_A\,A - \tilde c\,\bigr),
    && \text{(always observed)} \label{eq:obs-sleep}\\
  S(t) &\sim
    \Ncal\!\bigl(\,S_{\mathrm{base}} + k_F\,F - k_A^S\,A
                  + \beta_C^S\,C(t),\;
                  \sigma_S^2\bigr),
    && \text{(wake-gated)} \label{eq:obs-stress}\\
  \log(\mathrm{steps}(t)+1) &\sim
    \Ncal\!\bigl(\,\mu_0^{\mathrm{step}} + \beta_B^{\mathrm{step}}\,B
                  - \beta_F^{\mathrm{step}}\,F + \beta_A^{\mathrm{step}}\,A
                  + \beta_C^{\mathrm{step}}\,C(t),\;
                  \sigma_{\mathrm{step}}^2\bigr),
    && \text{(wake-gated)} \label{eq:obs-steps}
\end{align}
where $C(t) = \cos(2\pi t + \phi)$ is a fixed deterministic circadian
covariate (not a latent variable). ``Sleep-gated'' channels are emitted
only when the sleep label of \eqref{eq:obs-sleep} is one; ``wake-gated''
channels only when it is zero. The collection of observation-model
parameters $(\mathrm{HR}_{\mathrm{base}},\kappa_B^{\mathrm{HR}},\dots,\sigma_{\mathrm{step}})$
brings the full static parameter vector $\thetadyn$ for the joint
state-space model to roughly $30$ scalar entries. We will write
$y_{1:n}$ for the multi-channel observation record up to time $t_n$ and
$p(y_{1:n}\mid \thetadyn)$ for its marginal likelihood under the model.

\subsection{Two outstanding analyses}\label{sec:fsa-remarks}

The model defined by
\eqref{eq:fsa-mu}--\eqref{eq:fsa-A} and
\eqref{eq:obs-hr}--\eqref{eq:obs-steps} has been validated empirically by
the experiments reported in Section~\ref{sec:results}, but two analyses
that we would expect to be present in a complete treatment are
\emph{not yet} part of the project. We flag them here, before any of the
inference machinery is built, so that the reader does not mistake their
absence for completeness.

\begin{remark}[Fisher information rank identifiability --- outstanding]\label{rem:fim}
For a parameter vector $\theta \in \R^d$ and observations $y$ from the
model, recall that the Fisher information matrix is
\begin{equation}
  \Ical(\theta) \;=\;
    \E_{y\sim p(\cdot\mid\theta)}
    \!\Bigl[\,\nabla_\theta \log p(y\mid\theta)\,
              \nabla_\theta \log p(y\mid\theta)^{\!\top}\Bigr]. \label{eq:fim}
\end{equation}
Local identifiability of $\theta$ at a point $\theta^*$ requires that
$\Ical(\theta^*)$ have full rank $d$; rank-deficient directions
correspond to one-parameter families of models that produce
indistinguishable observation distributions. For FSA-v2 with $d \approx
30$, this rank check has \emph{not been carried out}. A future revision
of these notes should evaluate $\Ical(\theta^*)$ numerically -- for
instance via the score function obtained by automatic differentiation of
the model code -- at one or more representative truth points, and either
verify $\rank \Ical(\theta^*)=d$ or enumerate the rank-deficient
directions and their physical interpretation. Practical consequences are
discussed in Section~\ref{sec:discussion-fim}.
\end{remark}

\begin{remark}[Lyapunov stability of the FSA-v2 SDE --- outstanding]\label{rem:lyap}
A stability analysis for FSA-v2 would proceed in two stages, neither of
which has yet been carried out:
\begin{enumerate}
  \item[(a)] \emph{Deterministic skeleton.} For the zero-noise system
    under a constant control $\Phi \equiv \Phi_{\mathrm{const}}$, identify
    the equilibria $(B^*,F^*,A^*)(\Phi_{\mathrm{const}})$, compute the drift
    Jacobian, and check the eigenvalue signs (Lyapunov's first method).
    The case $A=0$ is always an equilibrium of \eqref{eq:fsa-A}; whether
    it is stable depends on the sign of $\mu(B^*,F^*)$. The Stuart--Landau
    bifurcation in $A$ implies that for some
    $\Phi_{\mathrm{const}}$ the deterministic skeleton has an additional
    stable limit cycle of radius $\sqrt{\mu(B^*,F^*)/\eta}$.
  \item[(b)] \emph{Stochastic stability.} Find a Lyapunov function
    $V:\R^3_+ \to \R_+$ for the full SDE such that the infinitesimal
    generator $\Lcal$ satisfies a Foster--Lyapunov drift condition
    $\Lcal V(x) \le -\alpha V(x) + c$ on the operating domain. Such an
    inequality would imply exponential ergodicity and the existence of a
    unique invariant distribution.
\end{enumerate}
Neither analysis has yet been carried out for FSA-v2; both are flagged
as required next steps in Section~\ref{sec:discussion-lyap}.
\end{remark}
