\section{The bootstrap particle filter} \label{sec:pf}

This section presents the inner Bayesian filtering layer. For a fixed
parameter vector $\thetadyn$, the particle filter estimates the
sequence of conditional state distributions
$\pi_n(\dd x_n) := p(\dd x_n \mid Y_{1:n}, \thetadyn)$ and, as a
by-product, returns an unbiased estimator of the marginal observation
likelihood $p(Y_{1:T} \mid \thetadyn)$. That estimator is the hook
that makes the outer SMC$^2$ of Section~\ref{sec:smc2} possible.

\subsection{The Bayesian filtering recursion}

Under the SDE--observation model of Section~\ref{sec:setting}, with
$\thetadyn$ fixed, the filtering distribution satisfies the
prediction--update recursion
\begin{align}
\pi_{n \mid n-1}(\dd x_n)
   &= \int p(\dd x_n \mid x_{n-1}, U_n, \thetadyn)\,
            \pi_{n-1}(\dd x_{n-1}),
   \label{eq:filter-predict} \\
\pi_n(\dd x_n)
   &\propto p(Y_n \mid x_n, C_n, \thetadyn)\;
            \pi_{n \mid n-1}(\dd x_n).
   \label{eq:filter-update}
\end{align}
Equations~\eqref{eq:filter-predict}--\eqref{eq:filter-update} have
closed form only when both densities are linear-Gaussian (the Kalman
case). In the general nonlinear non-Gaussian setting they do not, and
a Monte Carlo approximation is used.

\subsection{Sequential importance resampling (SIR)}

The bootstrap particle filter (Gordon, Salmond and Smith, 1993; Doucet
and Johansen, 2011) approximates $\pi_n$ by a weighted empirical
measure
\begin{equation}
\pi_n^N(\dd x_n) \;=\; \sum_{i=1}^{N} W_n^{(i)}\,
                        \delta_{X_n^{(i)}}(\dd x_n),
\qquad \sum_{i=1}^{N} W_n^{(i)} = 1,
\end{equation}
on a population of $N$ \emph{state particles} $\{X_n^{(i)}\}$. Each
particle carries a normalised importance weight $W_n^{(i)}$. The
recursion has three substeps per step: propagate, reweight, resample.
This is summarised in Algorithm~\ref{alg:sir}.

\begin{algorithm}[H]
\caption{Bootstrap SIR for a fixed $\thetadyn$}
\label{alg:sir}
\begin{algorithmic}[1]
\State Sample $X_0^{(i)} \sim p(\dd x_0 \mid \thetadyn)$ for
       $i = 1, \dots, N$; set $W_0^{(i)} = 1/N$.
\For{$n = 1, 2, \ldots$}
  \State \textbf{Propagate.} For each $i$, draw
         $X_n^{(i)} \sim p(\,\cdot \mid X_{n-1}^{(i)},\, U_n,\, \thetadyn)$
         by one Euler--Maruyama step of \eqref{eq:em}.
  \State \textbf{Reweight.} Set
         $\tilde W_n^{(i)} = W_{n-1}^{(i)}\,
            p(Y_n \mid X_n^{(i)},\, C_n,\, \thetadyn)$ and normalise
         $W_n^{(i)} = \tilde W_n^{(i)} / \sum_j \tilde W_n^{(j)}$.
  \State \textbf{ESS check.} Compute $N_{\mathrm{eff}} =
         1/\sum_i (W_n^{(i)})^2$. If $N_{\mathrm{eff}} < N_{\rm thr}$,
         systematically resample $\{X_n^{(i)}\}$ by $\{W_n^{(i)}\}$ and
         reset $W_n^{(i)} \leftarrow 1/N$.
\EndFor
\end{algorithmic}
\end{algorithm}

\paragraph{Systematic resampling.} The framework uses the
\emph{systematic} variant: with $u_0 \sim \Ucal[0, 1)$, the resampled
indices are obtained by walking the cumulative weight $C(j) =
\sum_{i \le j} W_n^{(i)}$ at the offsets $u_k = (k + u_0)/N$ for
$k = 0, \ldots, N-1$. Systematic resampling is $O(N)$, has lower
Monte Carlo variance than multinomial resampling, and is deterministic
given $u_0$. Implementation:
\path{blackjax.smc.resampling.systematic}, called from
\path{smc2fc/filtering/gk_dpf_v3_lite.py}.

\subsection{Liu--West shrinkage} \label{sec:liu-west}

The textbook bootstrap filter of Algorithm~\ref{alg:sir} is, in
production, run \emph{coupled} to the outer SMC$^2$ over $\thetadyn$
(Section~\ref{sec:smc2}). At each tempering step the parameter cloud
is jittered by HMC, and the inner filter is re-run. Without
intervention, this coupled-sampler scheme suffers from
\emph{parameter--state cloud collapse}: the inner state cloud's
variance shrinks too quickly because the resampling is not aware of
the parameter degree of freedom.

The Liu--West (2001) shrinkage correction addresses this. Let $h$ be
a Silverman bandwidth on the stochastic state subspace and define the
shrinkage coefficient $a = \sqrt{1 - h^2}$. After resampling, each
particle is moved toward the weighted state mean:
\begin{equation}
X_n^{(i)} \;\leftarrow\; a\, X_n^{(i)} \;+\; (1 - a)\, \mu_W,
\qquad \mu_W = \sum_i W_n^{(i)}\, X_n^{(i)}.
\label{eq:liu-west}
\end{equation}
The choice $a = \sqrt{1 - h^2}$ ensures the empirical variance of the
shrunk cloud equals the unshrunk variance plus the kernel-bandwidth
inflation, preserving the second moment exactly. In production the
bandwidth itself is \emph{ESS-scaled}: when the cloud is healthy
($N_{\rm eff} \approx N$) the bandwidth is near zero and the
shrinkage is a near-identity; when the cloud collapses
($N_{\rm eff} \to 1$) the bandwidth equals the Silverman default and
full shrinkage fires.

The continuous mapping $N_{\rm eff} \to h \to (\text{shrink amount})$
replaces a discontinuous threshold and is critical for
gradient-based outer-loop kernels: see the lengthy comment block in
\path{smc2fc/filtering/_gk_kernel.py:38--56} for the engineering history
(an MCLMC step-size tuner regression that was caused by the
discontinuity and fixed by ESS-scaled bandwidth).

\subsection{The unbiased marginal likelihood} \label{sec:pf-likelihood}

A key property of the bootstrap filter, which makes the SMC$^2$
algorithm of Section~\ref{sec:smc2} possible, is that it produces an
unbiased estimator of the marginal observation likelihood. With the
unnormalised weights $\tilde W_n^{(i)}$ from line~4 of
Algorithm~\ref{alg:sir}, define
\begin{equation}
\Lhat_N(\thetadyn) \;:=\; \prod_{n=1}^{T}
   \biggl(\frac{1}{N} \sum_{i=1}^{N} \tilde W_n^{(i)}\biggr).
\label{eq:unbiased-likelihood}
\end{equation}

\begin{proposition}[Unbiasedness; Del Moral, 2004] \label{prop:unbiased}
For any $N \ge 1$ and any $T \ge 1$,
\(
\E\bigl[\Lhat_N(\thetadyn)\bigr] = p(Y_{1:T} \mid \thetadyn).
\)
\end{proposition}

A proof is in Del Moral (2004, Theorem 7.4.2). For the present notes
the significance is operational: it allows us to plug
$\Lhat_N(\thetadyn)$ in place of the (intractable) exact likelihood
$p(Y_{1:T} \mid \thetadyn)$ inside any sampler over $\thetadyn$ and
obtain a procedure that targets exactly the right posterior. This is
precisely what SMC$^2$ does (Section~\ref{sec:smc2}).

\subsection{Implementation pointer}

The compile-once entry point is
\path{smc2fc.filtering.gk_dpf_v3_lite.make_gk_dpf_v3_lite_log_density_compileonce}.
Given an \texttt{EstimationModel} and inner-PF / OT-rescue
hyperparameters, it returns a JIT-compiled callable
\texttt{log\_density(u, grid\_obs, fixed\_init\_state, w\_start, key0)}
that runs the full bootstrap filter for a given parameter vector
$u = \thetadyn$ and observation buffer, returning $\log \Lhat_N(u)$
plus a small state dict.

The Gaussian-kernel resampling primitives that implement the
ESS-scaled Liu--West shrinkage of Section~\ref{sec:liu-west} live in
\path{smc2fc/filtering/_gk_kernel.py} (functions
\texttt{smooth\_resample\_ess\_scaled\_lw} and friends). The optional
optimal-transport rescue invoked inside the same scan is the subject
of Section~\ref{sec:ot-rescue}.
