\section{Code-API quick reference} \label{app:code-api}

A one-line description of every public function and class in the
\path{smc2fc/} package, organised by sub-package. The intent is the
``which file is X in?'' lookup --- not a substitute for reading the
docstrings.

\subsection{\texorpdfstring{\texttt{smc2fc/core/}}{smc2fc/core/}}

\begin{description}\small
  \item[\texttt{config.SMCConfig}] Master configuration dataclass for
        outer SMC$^2$, inner PF, OT-rescue, and bridge. See
        Section~\ref{sec:config}.
  \item[\texttt{config.RollingConfig}] Window/stride/dt configuration
        for rolling-window framing.
  \item[\texttt{config.MissingDataConfig}] Default model of
        observation corruption (rest days, dropout, broken-watch
        gaps) used in stress tests.
  \item[\texttt{jax\_native\_smc.run\_smc\_window\_native}]
        Cold-start tempered SMC$^2$ for one observation window.
        JAX-native, compile-once. Production entry point for the
        first window.
  \item[\texttt{jax\_native\_smc.run\_smc\_window\_bridge\_native}]
        Warm-start tempered SMC$^2$ for one observation window using
        the SF/BW bridge initialisation. Production entry point for
        windows $n \ge 2$ when \texttt{bridge\_type =
        "schrodinger\_follmer"}.
  \item[\texttt{jax\_native\_smc.run\_tempered\_chain}] The on-device
        tempered loop itself (single-window, single
        \texttt{lax.while\_loop}). Internal but exposed.
  \item[\texttt{jax\_native\_smc.TemperedState}] \texttt{NamedTuple}
        carrying particles, log-weights, current $\lambda$, and HMC
        kernel state across iterations of the on-device loop.
  \item[\texttt{jax\_native\_smc\_nuts.*}] NUTS variant of the above.
        Documented dead-end; see Section~\ref{sec:nuts-failed}. Not
        for production use.
  \item[\texttt{tempered\_smc.run\_smc\_window}] BlackJAX-wrapped
        cold-start equivalent of
        \texttt{run\_smc\_window\_native}. Slower, easier to debug.
  \item[\texttt{tempered\_smc.run\_smc\_window\_bridge}] BlackJAX-
        wrapped Gaussian-bridge warm-start. The
        \texttt{bridge\_type = "gaussian"} path lives here; the
        JAX-native side currently only implements the SF variant.
  \item[\texttt{mass\_matrix.estimate\_mass\_matrix}] Diagonal mass-
        matrix estimator from a particle cloud (variance with a
        small regularisation floor). Re-called per tempering level.
  \item[\texttt{sampling.sample\_from\_prior}] Convenience prior
        sampler in unconstrained space, used at cold start.
  \item[\texttt{sf\_bridge.bures\_wasserstein\_map}] BW transport
        matrix between two PSD covariances.
  \item[\texttt{sf\_bridge.bw\_geodesic}] Closed-form Bures--
        Wasserstein geodesic between two Gaussians, parameterised by
        $t \in [0, 1]$.
  \item[\texttt{sf\_bridge.estimate\_target\_gaussian}] Single-step
        importance-sampling fit of $q_1$ (the new posterior) given
        $q_0$ (the previous posterior).
  \item[\texttt{sf\_bridge.estimate\_target\_gaussian\_annealed}]
        $K$-stage tempered SMC fit of $q_1$ with random-walk
        Metropolis. Activated by \texttt{sf\_q1\_mode = "annealed"}.
  \item[\texttt{sf\_bridge.fit\_sf\_base}] Top-level entry point for
        the SF/BW bridge: returns the blended Gaussian to seed the
        next window.
  \item[\texttt{sf\_bridge.estimate\_fim\_hessian}] Diagonal Fisher-
        information estimate from the new-window log-likelihood
        Hessian; used by the info-aware blending mode.
  \item[\texttt{sf\_bridge.per\_eigenvector\_blend} /
        \texttt{info\_aware\_bridge\_mean}] Per-direction blending
        between $m_0$ and $m_1$ keyed off the FIM eigenvalues.
        Activated by \texttt{sf\_info\_aware = True}.
\end{description}

\subsection{\texorpdfstring{\texttt{smc2fc/filtering/}}{smc2fc/filtering/}}

\begin{description}\small
  \item[\texttt{gk\_dpf\_v3\_lite.make\_gk\_dpf\_v3\_lite\_log\_density\_compileonce}]
        Build the closed-over inner-PF log-likelihood for a single
        observation window. The compile-once variant the JAX-native
        backend consumes.
  \item[\texttt{gk\_dpf\_v3\_lite.make\_gk\_dpf\_v3\_lite\_log\_density}]
        Per-call variant; closes over data values rather than just
        shapes. Used by the BlackJAX-wrapped backend.
  \item[\texttt{\_gk\_kernel.compute\_ess}] Effective sample size from
        log-weights.
  \item[\texttt{\_gk\_kernel.silverman\_bandwidth}] Silverman's rule
        bandwidth for the smoothing kernel.
  \item[\texttt{\_gk\_kernel.log\_kernel\_matrix}] Log of the
        Gaussian kernel matrix between particles, used by the
        Liu--West smoothing step.
  \item[\texttt{\_gk\_kernel.smooth\_resample\_basic}] Plain
        systematic resample with Liu--West shrinkage and Gaussian
        smoothing.
  \item[\texttt{\_gk\_kernel.smooth\_resample\_ess\_scaled}] As
        above, with bandwidth scaled by ESS to avoid over-smoothing
        when ESS is high.
  \item[\texttt{\_gk\_kernel.smooth\_resample}] Adaptive wrapper that
        dispatches to one of the variants above depending on ESS.
  \item[\texttt{\_gk\_kernel.smooth\_resample\_ess\_scaled\_lw}]
        Log-weighted variant for use inside the bootstrap filter
        likelihood evaluation.
  \item[\texttt{transport\_kernel.compute\_kernel\_factor}] Nystr\"om
        low-rank kernel factor between $N$ particles and $r$ random
        anchors; the $K_{NR}$ matrix in
        Section~\ref{sec:ot-rescue}.
  \item[\texttt{transport\_kernel.factor\_matvec}] Mat-vec via the
        Nystr\"om factorisation; $O(Nr)$ instead of $O(N^2)$.
  \item[\texttt{transport\_kernel.factor\_matvec\_batch}] Batched
        version of the same.
  \item[\texttt{sinkhorn.sinkhorn\_scalings}] Sinkhorn iterations on
        the low-rank-kernel-factored transport problem, returning the
        $(u, v)$ scaling vectors.
  \item[\texttt{project.barycentric\_projection}] Barycentric
        projection of a source cloud onto a target cloud given
        Sinkhorn scalings $(u, v)$ and the kernel factor.
  \item[\texttt{resample.ot\_resample\_lr}] Top-level OT-rescue:
        Sinkhorn $\to$ projection $\to$ sigmoid blend with the
        systematic-resample output.
\end{description}

\subsection{\texorpdfstring{\texttt{smc2fc/control/}}{smc2fc/control/}}

\begin{description}\small
  \item[\texttt{control\_spec.ControlSpec}] Dataclass describing the
        controller's task: cost callable, schedule decoder, horizon
        length, initial state estimate, reference dynamics
        parameters.
  \item[\texttt{config.SMCControlConfig}] Configuration dataclass for
        the controller's tempered-SMC$^2$ loop (counterpart to
        \texttt{SMCConfig} on the inference side).
  \item[\texttt{tempered\_smc\_loop.run\_tempered\_smc\_loop\_native}]
        JAX-native control SMC$^2$ loop: tempers the prior toward
        the cost-as-likelihood and returns the posterior cloud
        (mean = recommended $\bar\theta_{\rm ctrl}$).
  \item[\texttt{tempered\_smc\_loop.run\_tempered\_smc\_loop}]
        BlackJAX-wrapped reference equivalent.
  \item[\texttt{rbf\_schedules.RBFSchedule}] Gaussian-RBF schedule
        decoder \eqref{eq:schedule-decoder}: maps a coefficient
        vector to a $U(t)$ trajectory.
  \item[\texttt{calibration.calibrate\_beta\_max}] Choose the high-
        $\lambda$ inverse temperature from the cost spread of the
        prior cloud.
  \item[\texttt{calibration.build\_crn\_noise\_grids}] Pre-sample
        common-random-number diffusion shocks for use across the
        tempering schedule.
  \item[\texttt{diagnostics.plot\_cost\_histogram} /
        \texttt{plot\_schedule\_comparison} /
        \texttt{plot\_trajectories} / \texttt{print\_smc\_step} /
        \texttt{evaluate\_gates}] Post-hoc inspection helpers; not on
        the run-time critical path.
  \item[\texttt{lqg.controller.LQGSpec} / \texttt{LQGController} /
        \texttt{build\_lqg\_open\_loop\_schedule}] LQG baseline with
        a finite-horizon Riccati solver. Provides a sanity-check
        comparator to the tempered-SMC$^2$ controller.
  \item[\texttt{lqg.linearize.linearize\_drift\_at}] Local
        linearisation of the SDE drift at an operating point; used
        by the LQG baseline.
  \item[\texttt{lqg.riccati.solve\_riccati\_backward} /
        \texttt{compute\_lqr\_gain}] Backward Riccati integration and
        LQR gain extraction.
\end{description}

\subsection{\texorpdfstring{\texttt{smc2fc/simulator/}}{smc2fc/simulator/}}

\begin{description}\small
  \item[\texttt{sde\_model.StateSpec}] Description of a single SDE
        state component (name, prior, bounds).
  \item[\texttt{sde\_model.ChannelSpec}] Description of a single
        observation channel (which states it sees, noise model).
  \item[\texttt{sde\_model.SDEModel}] The user-supplied model
        interface: drift, diffusion, observation kernel, parameter
        priors. Section~\ref{sec:setting} expands on this.
  \item[\texttt{sde\_solver\_diffrax.solve\_deterministic\_jax}]
        Deterministic-mean trajectory via diffrax (used for warm-
        starts and diagnostics).
  \item[\texttt{sde\_solver\_diffrax.solve\_sde\_jax}] Stochastic SDE
        solve via diffrax with optional control schedule.
  \item[\texttt{sde\_observations.generate\_all\_channels}] Sample
        per-channel observations from a trajectory.
\end{description}

\subsection{\texorpdfstring{\texttt{smc2fc/transforms/} and top-level}{smc2fc/transforms/ and top-level}}

\begin{description}\small
  \item[\texttt{transforms.unconstrained.build\_transform\_arrays}]
        Build the bound/scale arrays for the unconstrained-space
        reparameterisation.
  \item[\texttt{transforms.unconstrained.constrained\_to\_unconstrained}
        / \texttt{unconstrained\_to\_constrained}] Bijection between
        the user-facing constrained $\thetadyn$ and the HMC-friendly
        unconstrained $u$.
  \item[\texttt{transforms.unconstrained.log\_prior\_unconstrained}]
        Log-prior in unconstrained space (with Jacobian correction).
  \item[\texttt{transforms.unconstrained.split\_theta}] Slice an
        unconstrained vector into dynamics and control halves.
  \item[\texttt{estimation\_model.EstimationModel}] User-side
        inference interface (parameter list, bounds, log-prior,
        link to \texttt{SDEModel}).
\end{description}
