\section{Results}\label{sec:results}

The two headline empirical results of the project are an open-loop
benchmark on the canonical $84$-day horizon (one chronic time
constant of $42$ days, doubled) and the closed-loop fourteen-day
rolling controller. Both compare the SMC${}^2$ schedule to the
canonical constant baseline $\Phi \equiv 1.0$. Headline numbers are
summarised in Table~\ref{tab:results}.

\begin{table}[h]
\centering
\small
\begin{tabular}{@{}lccccc@{}}
\toprule
Benchmark & Horizon & $\bar A$ (SMC${}^2$) & $\bar A$ (baseline) & gain & F-violation \\
\midrule
Open-loop, Stage D     & $84$ d & $0.645$  & $0.503$  & $+28\%$ & $1.28\%$ \\
Closed-loop MPC, Stage E5 & $14$ d & $0.0964$ & $0.0823$ & $+17\%$ & $0.00\%$ \\
\bottomrule
\end{tabular}
\caption{Headline empirical results. $\bar A$ denotes the mean
autonomic amplitude $\frac{1}{T}\int_0^T A(t)\,\dd t$ averaged over
Monte Carlo trajectories. Sources:
\texttt{version\_1/outputs/fsa\_high\_res/RESULT.md} (Stage D),
\texttt{version\_2/outputs/fsa\_high\_res/RESULT.md} (Stage E5).}
\label{tab:results}
\end{table}

\subsection{Stage D: open-loop schedule on the $84$-day horizon}

In the open-loop benchmark, the dynamics parameters $\thetadyn$ are
treated as known at their truth values, the initial state is fixed
at $(B_0,F_0,A_0) = (0.05,0.30,0.10)$ (a de-trained athlete with
high residual fatigue), and the SMC${}^2$ control of
Section~\ref{sec:control} is run \emph{once} over the full $84$-day
horizon. The output is a single posterior-mean schedule
$\Phi(\,\cdot\,;\bar\theta)$.

The optimised schedule has mean $\Phi \approx 1.96$ over the horizon
and a front-loaded periodisation profile (heavy training for the
first $\sim 60$ days, taper thereafter). Mean amplitude
$\bar A = 0.645$ exceeds the constant baseline $\bar A = 0.503$ by
$28\%$, exceeds the sedentary reference $\bar A = 0.209$ by a factor
of $3.08$, and the soft fatigue barrier is violated only $1.28\%$
of the time --- well within the $5\%$ acceptance gate of
Section~\ref{sec:mpc-feasibility}. All four acceptance gates
(mean-A matches baseline within $3\%$, mean-A $\ge 1.40\times$
sedentary, mean-$\Phi \in [0.5,2.5]$, F-violation $\le 5\%$) pass.

\subsection{Stage E5: closed-loop MPC on the $14$-day macrocycle}

In the closed-loop benchmark the dynamics parameters are no longer
known to the controller. Algorithm~\ref{alg:fsa-mpc} is run on a
$14$-day record of four-channel observations, with a one-day filter
window, twelve-hour stride, and $27$ replans. The plant is the
\texttt{StepwisePlant} of Section~\ref{sec:mpc-code}.

Mean amplitude under the closed-loop policy is $\bar A = 0.0964$,
against $\bar A = 0.0823$ for the constant baseline
$\Phi \equiv 1.0$ on the same $14$-day horizon. (The shorter horizon
explains the smaller absolute numbers compared to Stage D: the
integral $\int_0^T A(t)\,\dd t$ has not yet converged to the
asymptotic mean.) The $+17\%$ gain is achieved with a remarkable
schedule shape: the controller commits a \emph{rest-leaning} $\Phi$
in the range $0.25$--$0.31$ across all $27$ strides, against the
canonical baseline of $1.0$. Physiologically, this is the controller
exploiting the fact that the simulated athlete starts with high
residual fatigue ($F_0 = 0.30$): the optimal short-horizon strategy
is to clear fatigue rather than to add training stimulus, which
unlocks the Stuart--Landau growth-rate $\mu(B,F)$ and lets $A$ rise
from $0.10$ to $\sim 0.20$ over the macrocycle.

The fatigue ceiling is never violated ($0.00\%$). Wall-clock cost on
a single GPU is $41$ minutes for the full closed loop. The
closed-loop policy is robust despite imperfect filter coverage --
on this record only $3$ of $26$ estimable parameters meet a strict
$\ge 5/6$ coverage gate at the window level -- consistent with the
flat-cost-surface argument of Section~\ref{sec:mpc-robustness}.

\subsection{Stage~I: cross-grid consistency at \texorpdfstring{$h = 1$ h}{h=1h}}\label{sec:results-stage-i}

The $1$-hour outer step recommended in
Section~\ref{sec:num-stab-recommend} has been implemented as
Stage~I, behind the \texttt{FSA\_STEP\_MINUTES} environment
variable and the \texttt{--step-minutes} CLI flag of the bench
drivers. The controller's inner sub-step is automatically scaled
with the outer step ($n_{\mathrm{sub}} = 4$ at $96$ bins/day,
$n_{\mathrm{sub}} = 1$ at $24$ bins/day) so the speedup is
preserved through the controller cost MC.

The four sanity tests that gate the Stage~I refactor all pass: the
grid constants for $\mathrm{step\_minutes} \in \{15,30,60\}$ are
internally consistent, the circadian channel $C(t)$ at hour-aligned
bins agrees across grids, the sleep gating preserves an $8$-hour
sleep window on both grids, and the plant's one-day means agree
across the two grids to $1.64\%$ relative error. End-to-end speedup
is roughly $4\times$, projecting the $T = 84$ closed-loop run from
$\sim\!14$ GPU-hours down to $\sim\!3.5$ GPU-hours on an RTX~5090.
The remaining cross-grid validation gate at $T = 14$ for the
SMC${}^2$ MPC will close once the corresponding $h = 1$ h run lands;
this is work in progress on the math laptop.

\medskip
\noindent
A trajectory plot of $A(t)$ under the SMC${}^2$ MPC closed-loop
policy and the constant baseline $\Phi \equiv 1$, together with the
discovered $\Phi$ schedule, can be regenerated from the artefacts in
\texttt{version\_2/outputs/fsa\_high\_res/} by re-running
\texttt{tools/bench\_smc\_full\_mpc\_fsa.py}.

\medskip
\noindent
A linear-quadratic-Gaussian alternative to the SMC${}^2$ MPC was
also examined and benchmarked across the same horizons; the full
account is deferred to Appendix~\ref{app:lqg}. The verdict is brief:
the cubic Stuart--Landau non-linearity at the heart of FSA-v2 cannot
be captured by any LQG variant, and the LQG controller does not
match the SMC${}^2$ MPC's performance on this plant.
